Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-05-31 analysis/paper/paper.Rmd
Adjusting priors

All regression coefficients are assumed to be Gaussian distributed, with
priors derived from the expected values and 95% confidence intervals
estimated by Johnson et al (2020) (Table \@ref(tab:priors)). We
calibrated the prior distributions so that the 95% prior prediction
intervals for survival and recruitment from the Bayesian model match the
range between the 2.5% and 97.5% quantiles of 1000 simulated survival
and recruitment trajectories (Supplement C, Figure
\@ref(fig:examplePredictions)).

@@

knitr::kable(tbl,caption="Prior means and standard errors of Gaussian distributed survival and reproduction model parameters. Standard errors are calibrated so that the 95\\% prior prediction intervals for survival and recruitment from the Bayesian model match the range between the 2.5\\% and 97.5\\% quantiles of 1000 simulated survival
and recruitment trajectories (Supplement C).")
#TO DO: indicate which of these are from Johnson, and which are calibrated.
```

>>>>

All regression coefficients are assumed to be Gaussian distributed, with
prior means from Johnson et al's (2020) regression model estimates (Table \@ref(tab:priors)). We
calibrated the standard deviations of the priors so that the 95% prior predictive
intervals for survival and recruitment from the Bayesian model match the
or exceed the 95% predictive intervals from Johnson et al's regression model (Supplement C, Figure
\@ref(fig:examplePredictions)).

@@

knitr::kable(tbl,caption="Prior means and standard deviations of Gaussian distributed survival and reproduction model parameters. Standard deviations are calibrated so that the 95\\% prior predictive intervals for survival and recruitment from the Bayesian model match the
or exceed the 95% predictive intervals from Johnson et al's regression model (Supplement C).")
```

### 2024-05-31 analysis/paper/supplement.Rmd
Adjusting priors

Prior parameter values should be set so that, given no monitoring data, the prior survival distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance. 
This condition is met when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (sSigmaMean & sSigmaSD, $\sigma^2_S$ in Table 2) (Figs \@ref(fig:PriorKSDists) and \@ref(fig:PriorAllFits)). Priors should also be vague enough so that, given intensive monitoring of populations where survival is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 60 collars per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local survival, rather than priors from the national model. Setting the mean and standard deviations of the prior distribution of the random effect of year on survival to moderately low values (sSigmaMean=0.034784, sSigmaSD=0.03, $\sigma^2_S$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDists), \@ref(fig:PriorAllFits) and \@ref(fig:PriorLowFits)).

Similarly, prior parameter values should be set so that, given no monitoring data, the prior recruitment distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance. This condition is met when the multiplier on the prior standard deviation of the recruitment intercept (rIntSEMod: $\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (rSigmaMean & rSigmaSD, $\sigma^2_R$ in Table 2) (Figs \@ref(fig:PriorKSDistsRec) and \@ref(fig:PriorAllFitsRec)). Priors should also be vague enough so that, given intensive monitoring of populations where recruitment is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 9*60 cows per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local recruitment, rather than priors from the national model. Setting the mean and standard deviations of the prior distribution of the random effect of year on recruitment to moderately low values (rSigmaMean=0.23, rSigmaSD=0.22, $\sigma^2_R$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDistsRec), \@ref(fig:PriorAllFitsRec) and \@ref(fig:PriorHighFitsRec)).

```{r priorCalibration, cache=TRUE,echo=FALSE, message=FALSE,results='hide'}
if(doPriorCalibration){

  source(here::here("analysis/scripts/calibratePriors.R"))
}else{
  KS<-read.csv(here::here("tabs/SurvKSAll.csv"))
  KSRec<-read.csv(here::here("tabs/RecKSAll.csv"))

}
  #library(ggplot2);fig_widths=data.frame(two=7)
  KS=subset(KS,scn=="All")
  png(here::here("figs/PriorKSDists.png"),
    height = 3, width = fig_widths["two"], units = "in",res=600)
    base = ggplot(KS,aes(x=sIntSEMod,y=KSDistance,color=sSigmaMean))+
    geom_point()+facet_wrap(~sSigmaSD,labeller="label_both")+theme_bw()+ylab("Kolmogorov-Smirnov Distance")+xlab("sIntSEMod: multiplier on SD of survival intercept")
    print(base)
  dev.off()

  KSRec=subset(KSRec,scn=="All")
  png(here::here("figs/PriorKSDistsRec.png"),
    height = 3, width = fig_widths["two"], units = "in",res=600)
    base = ggplot(KSRec,aes(x=rIntSEMod,y=KSDistance,color=rSigmaMean))+
    geom_point()+facet_wrap(~rSigmaSD,labeller="label_both")+theme_bw()+ylab("Kolmogorov-Smirnov Distance")+xlab("rIntSEMod: multiplier on SD of recruitment intercept")
    print(base)
  dev.off()
```

```{r PriorKSDists, fig.cap="Similarity of the prior survival distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022]. The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (sSigmaMean & sSigmaSD, $\\sigma^2_S$ in Table 2). Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
knitr::include_graphics(here::here("figs/PriorKSDists.png"))

```{r PriorAllFits, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, the standard deviation of the random effect prior (sSigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}

knitr::include_graphics(here::here("figs/SurvInterceptAllsSigmaSD0.03.png"))
```

```{r PriorLowFits, fig.cap="Posterior predictive survival distributions from the Bayesian model ('local') assuming intensive monitoring (15 years of 60 collars per year) of populations from the 2.5 \\% quantile of the national model. In these examples, the standard deviation of the random effect prior (sSigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and  anthropogenic disturbance is 0. Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates. When the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is too low (<5) the insufficiently vague priors substantially influence the posterior distributions, yielding poor local parameter estimates.", cache=TRUE}

knitr::include_graphics(here::here("figs/SurvInterceptLowsSigmaSD0.03.png"))
```

```{r PriorAllFitsAnthro90, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, anthropogenic disturbance is 90\\%. When the multiplier on the prior standard deviation of the survival slope (sSlopeSEMod: $\\beta^S_a$ in Table 2) is 3, the posterior survival distribution is similar to the distribution of simulated outcomes from the national model. Bands show 95\\% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.", cache=TRUE}
knitr::include_graphics(here::here("figs/SurvSlopeAll.png"))
```

```{r PriorHighFitsAnthro90, fig.cap="Posterior predictive survival distributions from the Bayesian model, assuming intensive monitoring (15 years of 60 collars per year) of populations from the 97.5 \\% quantile of the national model. In these examples, anthropogenic disturbance is 90\\%, and the effect of anthropogenic disturbance varies from none (sSlopeMod=0) through expected (sSlopeMod=1) to double the expected slope (sSlopeMod=2). All considered multipliers on the prior standard deviation of the survival slope (sSlopeSEMod: $\\beta^S_a$ in Table 2) are sufficiently vague to reflect local variation among populations. Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates.", cache=TRUE}
knitr::include_graphics(here::here("figs/SurvSlopeHigh.png"))
```

```{r PriorKSDistsRec, fig.cap="Similarity of the prior recruitment distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022]. The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (rIntSEMod: $\\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (rSigmaMean & rSigmaSD, $\\sigma^2_R$ in Table 2). Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
knitr::include_graphics(here::here("figs/PriorKSDistsRec.png"))
```

```{r PriorAllFitsRec, fig.cap="Differences between prior predictive recruitment distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, the standard deviation of the random effect prior (rSigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}
knitr::include_graphics(here::here("figs/RecInterceptAllrSigmaSD0.22.png"))
```

```{r PriorHighFitsRec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model ('local') assuming intensive monitoring (9*60 cows per year for 15 years) of populations from the 97.5% quantile of the national model.  In these examples, the standard deviation the random effect prior (rSigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and  anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals, black dots show true recruitment rates from the simulated population, and open triangles show observed recruitment rates.", cache=TRUE}
knitr::include_graphics(here::here("figs/RecInterceptHighrSigmaSD0.22.png"))
```

```{r PriorAllFitsAnthro90Rec, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, anthropogenic disturbance is 90\\%. When the multiplier on the prior standard deviation of the recruitment slope (rSlopeSEMod: $\\beta^R_a$ in Table 2) is 4, the posterior survival distribution is similar to the distribution of simulated outcomes from the national model. Bands show 95\\% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.", cache=TRUE}
knitr::include_graphics(here::here("figs/RecSlopeAll.png"))
```

```{r PriorHighFitsAnthro90Rec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model, assuming intensive monitoring (9*60 cows per year for 15 years) of populations from the 97.5 \\% quantile of the national model. In these examples, anthropogenic disturbance is 90\\%, and the effect of anthropogenic disturbance varies from none (rSlopeMod=0) through expected (rSlopeMod=1) to double the expected slope (rSlopeMod=2). All considered multipliers on the prior standard deviation of the recruitment slope >1 (rSlopeSEMod: $\\beta^R_a$ in Table 2) are sufficiently vague to reflect local variation among populations. Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates.", cache=TRUE}
knitr::include_graphics(here::here("figs/RecSlopeHigh.png"))
```

>>>>

We begin by setting random effects to zero, varying the prior standard deviations of the survival intercept (sIntSE is $\beta^S_0$ in Table 2) and recruitment intercept (rIntSE is $\beta^R_0$ in Table 2), and comparing the prior predictive intervals of the Bayesian model to the 95\% predictive intervals from Johnson et al.'s (2020) regression analysis in a case with no anthropogenic disturbance (top row of Fig \@ref(fig:InterceptsAndSlopes)). We set the prior standard deviation intercept parameters (sIntSE = 0.04, rIntSE = 0.25) so that the upper bounds of the Bayesian predictive intervals match the upper bounds of Johnson et al.'s predictive intervals.

We then vary the prior standard deviations of the slopes of the relationships between anthropogenic disturbance and demographic rates (sAnthroSlopeSE is $\beta^S_R$ and rAnthroSlopeSE is $\beta^R_R$ in Table 2), and compare the prior predictive intervals of the Bayesian model to the 95\% predictive intervals from Johnson et al.'s (2020) regression analysis in a case with 90\% anthropogenic disturbance (bottom row of Fig \@ref(fig:InterceptsAndSlopes)). We set the prior standard deviation of the slopes (sAnthroSlopeSE = 0.0005, rAnthroSlopeSE = 0.006) so that the upper bound of the Bayesian predictive interval does not substantially exceed the upper bound of Johnson et al.'s predictive interval.

The choice of priors for the random effects of year on survival and recruitment are less important because the random effects have no effect on expected population growth rate (Figs \@ref(fig:PriorAllFits) and \@ref(fig:PriorAllFitsR)). We set the hyperprior for the standard deviation of the survival random effect to vary uniformly between sSigmaMin = 0.00001 and sSigmaMax = 0.06 to exceed the variation among national model simulations by a moderate amount (Fig \@ref(fig:PriorAllFits)). We set the hyperprior for the standard deviation of the recruitment random effect to vary uniformly between rSigmaMin = 0.00001 and rSigmaMax = 0.2 to exceed the variation among national model simulations by a moderate amount (Fig \@ref(fig:PriorAllFitsR)).

Given intensive monitoring (15 years of 60 collars per year) of populations where demographic rates are higher or lower than the national average (from 97.5% and 2.5% quantiles of the national model), and the slope of the anthropogenic disturbance relationship differs from the national model, these priors are vague enough to allow posteriors to reflect local demographic rates (Figs \@ref(fig:SurvBoth) and \@ref(fig:RecBoth)).

```{r priorCalibration, cache=TRUE,echo=FALSE, message=FALSE,results='hide'}
if(doPriorCalibration){
  source(here::here("analysis/scripts/PriorsFromPIs.R"))
  source(here::here("analysis/scripts/calibratePriors.R"))
}
```

```{r InterceptsAndSlopes, fig.cap="Effect of variation in survival and recruitment prior standard deviations on the prior 95% predictive intervals and means of the Bayesian model (orange bands and lines) with no random effects (no interannual variation). Means and 95% predictive intervals from Johnson et al.'s 2020 regression model (black lines) summarize observed variation across the Canada." , cache=TRUE}
knitr::include_graphics(here::here("figs/InterceptsAndSlopes.png"))
```

```{r PriorAllFits, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('Bayesian') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}

knitr::include_graphics(here::here("figs/SurvInterceptAllsIntSE0.04.png"))
```

```{r PriorAllFitsR, fig.cap="Differences between prior predictive recruitment distributions from the Bayesian model ('Bayesian') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}

knitr::include_graphics(here::here("figs/RecInterceptAllrIntSE0.25.png"))
```

```{r SurvBoth, fig.cap="Posterior predicted survival from the Bayesian model fitted to 15 years of data, with 60 collars per year. Survival in these example populations is lower (sQuantile=0.025) or higher (sQuantile=0.975) than the national average, with no anthropogenic disturbance slope (sSlopeMod = 0), or an anthropogenic disturbance slope that is twice as steep as expected (sSlopeMod = 2). In these examples anthropogenic disturbance is 90%. Bands show 95% posterior predictive intervals, dots show true survival, and triangles show simulated observations.",  cache=TRUE}

knitr::include_graphics(here::here("figs/SurvBoth.png"))
```

```{r RecBoth, fig.cap="Posterior predicted recruitment from the Bayesian model fitted to 15 years of data, with 60 collars per year. Recruitment in these example populations is lower (rQuantile=0.025) or higher (rQuantile=0.975) than the national average, with no anthropogenic disturbance slope (rSlopeMod = 0), or an anthropogenic disturbance slope that is twice as steep as expected (rSlopeMod = 2). In these examples anthropogenic disturbance is 90%. Bands show 95% posterior predictive intervals, dots show recruitment, and triangles show simulated observations.",  cache=TRUE}

knitr::include_graphics(here::here("figs/RecBoth.png"))
```
